---
title: "EDS 213 Group Project - Group 04 "
author: "Cullen Molitor, Desik Somasundaram, Julia Parish, and Ryan Muunikhuis"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(metajam)
library(tidyverse)
library(here)
library(lubridate)
library(httr2)
library(purrr)
library(janitor)
library(XML)
library(zoo)
library(vegan)
```


Our group discovered the Scott Reef and Rowley Shoals Coral Bleaching Data dataset while searching the DataOne repository for “coral bleaching.” This particular dataset focuses on long-term monitoring data from 1994 at reef slope habitats off the coast of northwestern Australia. Data Source: https://search.dataone.org/view/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fmetadata%2Feml%2Fedi%2F952%2F1. The metadata includes the reef system, date range, taxonomy, habitat type, coral coverage, and dataset methods. 

```{r error =TRUE}
# assign data url to access coral data from DataOne then download
data_url <- "https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fedi%2F952%2F1%2Ff6212784c45d0a077f2c863868d22c4b"
# If data is akready up to date, this will throw an error
# error=T in header will move past this issue
download_d1_data(data_url, dir_name = "coral_data", path = ".")
```

```{r}
# assign data path
data_path <- here::here("coral_data")
```

```{r}
# Read in data with metajam
coral_list <- read_d1_files(data_path)
```

```{r}
# Pick out the data read in above and clean it up
corals <- coral_list$data %>% 
  janitor::clean_names() %>% 
  mutate(year_decimal = format(date_decimal(year_decimal), "%Y-%m-%d"),
         month = month(year_decimal),
         year = year(year_decimal)) %>% 
  rename(date = year_decimal)
```

```{r}
# Read in  ONI to be joined to coral data
oni <- read.table(
  "https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt",
  header = T) %>%
  dplyr::mutate(date = as.Date(ISOdate(YR, MON, 1)),
                date_start = as.Date(ISOdate(YR, MON, 1)),
                date_end = ceiling_date(date_start, "month")) %>%
  dplyr::rename(oni_anomaly = ANOM,
                month = MON,
                year = YR) %>% 
  dplyr::select(year, month, oni_anomaly, date_start, date_end) %>% 
  dplyr::mutate(roll_3_month_mean = zoo::rollmean(x = oni_anomaly, k = 3, fill = NA, align = "right")) %>% 
  filter(date_start > ymd("1994-09-01"))
```

```{r}
# Calculate the Simpson's diversity index 
# join corals and ONI data
coral_oni <- corals %>% 
  pivot_longer(cols = 5:18, names_to = "species", values_to = "cover") %>% 
  dplyr::group_by(system, reef, year, month, location) %>% 
  dplyr::summarise(simpson_index = vegan::diversity(cover, index = "simpson")) %>% 
  dplyr::left_join(oni %>% select(-date_start, -date_end)) %>% 
  dplyr::mutate(date = lubridate::make_date(year = year, month = month, day = 1))
```

```{r}
# Un-comment line below to write the updated csv file
# readr::write_csv(coral_oni, here("coral_diversity_oceanic_nino_index.csv"))
```

```{r fig.height=4, fig.width=9}
# Plot the coral diversity data with ONI color bar
ggplot() +
  ggplot2::geom_rect(
    data = oni,
    aes(xmin = date_start, xmax = date_end, ymin = -Inf, ymax = .3, fill = oni_anomaly)) +
  ggplot2::scale_fill_viridis_c(
    option = "plasma",
    guide = guide_colorbar(direction = "horizontal", title.position = "top",
                           order = 2, barheight = unit(.2, "cm"))) +
  scale_x_date(date_breaks = "2 year", date_labels = "%Y", expand = expansion(mult = c(0,0))) +
  geom_line(data = coral_oni, size = 1,
              aes(x = date, y = simpson_index, color = location)) +
  scale_color_viridis_d() +
  ggplot2::guides(color = guide_legend(order = 1)) +
  labs(fill = "Oceanic Ni\u00f1o Index", color = "Sites", 
       x = "Date", y = "Simpsons Diversity Index") +
  theme_minimal()
```

```{r}
# In the future think about an appropriate model and explore the relationship
# mod <- lm(simpson_index ~ roll_3_month_mean + location, data = coral_oni)
# summary(mod)
```

